Tutorial: Fitting log-Gaussian mixed-effects mixture models for keystroke data

### 2024-10-26 slides/sections/statisticalmodels.Rmd
After review update

```{r fig.width=5}
```
</div>

```{r fig.width=5}
linewidth <- .35
data <- data.frame(x = c(200, 1500))
word_mean <- 800
sent_mean <- 950
sd <- c(40, 60)
word_lbl <- "word initial"
sent_lbl <- "sentence initial"

@@

```{r fig.width=5}
ggplot(data = data, aes(x)) +
  stat_function(fun = dnorm, n = 101, args = list(mean = word_mean, sd = sd[1]), 
                linewidth = linewidth,
                aes(colour = "word initial")) +
  stat_function(fun = dnorm, n = 101, args = list(mean = sent_mean, sd = sd[2]), 
                linewidth = linewidth,
                aes(colour = "sentence initial"), show.legend = F) +
  geom_area(stat = "function", fun = dnorm, aes(fill = word_lbl), alpha = .25, 
            args = list(mean = word_mean, sd = sd[1])) +
  geom_area(stat = "function", fun = dnorm, aes(fill = sent_lbl), alpha = .25, 
            args = list(mean = sent_mean, sd = sd[2])) +
  geom_curve(aes(x = word_mean, xend = sent_mean, y = 0.01, yend = 0.00675),
             curvature = -0.2, 
             size = .25,
             arrow = arrow(length = unit(0.3, "cm"))) +
  annotate("text", x=(sent_mean+word_mean)/2, y=.0105, label=paste0("beta[diff]"), size = 4, parse=TRUE) +
  scale_color_manual(breaks = c(word_lbl, sent_lbl), values = ggthemes::colorblind_pal()(2)) +
  scale_fill_manual(breaks = c(word_lbl, sent_lbl), values = ggthemes::colorblind_pal()(2)) +
  scale_y_continuous(limits = c(0, .012)) +
  scale_x_continuous(labels = scales::comma, limits = c(650, 1150)) +
  labs(x = "Interkey intervals [in ms]", 
       colour = "Text location",
       fill = "Text location") +
  theme(axis.text.y = element_blank(),
        axis.title.y = element_blank(),
        axis.ticks.y = element_blank())
```
</div>

@@

$$
\text{iki}_i \sim\text{ } \text{log}\mathcal{N}(\beta_\text{textlocation[i]}, \sigma_{e_\text{textlocation[i]}}^2)
$$

@@

@@

@@

@@

@@

```{r fig.width=5}
ggplot(data, aes(x)) +
  stat_function(geom = "line", fun = plot_mix_comps,
                aes(colour = word_lbl),
                linewidth = linewidth,
                args = list(alpha, sigmae, lam = 1 - theta_simp)) +
  stat_function(geom = "line", fun = plot_mix_comps,
                aes(colour = sent_lbl),
                linewidth = linewidth,
                args = list(alpha, sigmae, lam = 1 - theta_conj)) +
  stat_function(geom = "line", fun = plot_mix_comps,
                aes(colour = word_lbl),
                linewidth = linewidth,
                args = list(alpha2_w, sigmae_p_w, lam = theta_simp)) +
  stat_function(geom = "line", fun = plot_mix_comps,
                aes(colour = sent_lbl),
                linewidth = linewidth,
                args = list(alpha2_s, sigmae_p_s, lam = theta_conj)) +
  geom_area(stat = "function", fun = plot_mix_comps, aes(fill = word_lbl), alpha = .25, 
            args = list(alpha, sigmae, lam = 1 - theta_simp)) +
  geom_area(stat = "function", fun = plot_mix_comps, aes(fill = sent_lbl), alpha = .15,
            args = list(alpha, sigmae, lam = 1 - theta_conj)) +
  geom_area(stat = "function", fun = plot_mix_comps, aes(fill = word_lbl), alpha = .25, 
            args = list(alpha2_w, sigmae_p_w, lam = theta_simp)) +
  geom_area(stat = "function", fun = plot_mix_comps, aes(fill = sent_lbl), alpha = .15,
            args = list(alpha2_s, sigmae_p_s, lam = theta_conj)) +
  geom_curve(aes(x = alpha2_w, 
                 xend = alpha2_s, y = 0.0008, yend = 0.0015),
             curvature = -0.2, 
             size = .15,
             arrow = arrow(length = unit(0.3, "cm"))) +
  annotate("text", 
           x = (alpha2_w + alpha2_s)/2,
           y=.0022, 
           label=paste0("theta[diff]", "~and~", "delta[diff]"), size = 4, parse=TRUE) +
  scale_color_manual(breaks = c(word_lbl, sent_lbl), values = ggthemes::colorblind_pal()(2)) +
  scale_fill_manual(breaks = c(word_lbl, sent_lbl), values = ggthemes::colorblind_pal()(2)) +
  scale_x_continuous(labels = scales::comma, 
                     limits = c(50, 1000), 
                     breaks = seq(0, 1500, 200)) +
  labs(x = "Interkey intervals [in ms]", 
       colour = "Text location",
       fill = "Text location") +
  theme(axis.text.y = element_blank(),
        axis.title.y = element_blank(),
        axis.ticks.y = element_blank())

$$
\text{iki}_{i} \sim\text{ } \theta_\text{location[i]} \times \text{log}\mathcal{N}(\beta + \delta_\text{location[i]}, \sigma_{e'_\text{location[i]}}^2) + \\
   (1 - \theta_\text{location[i]}) \times \text{log}\mathcal{N}(\beta, \sigma_{e_\text{location[i]}}^2)
$$

>>>>

```{r fig.width=5}
linewidth <- .35
data <- data.frame(x = c(200, 1500))
word_mean <- 800
sent_mean <- 950
sd <- c(40, 40)
word_lbl <- "word initial"
sent_lbl <- "sentence initial"

@@

```{r fig.width=5}
ggplot(data = data, aes(x)) +
  stat_function(fun = dnorm, n = 101, args = list(mean = word_mean, sd = sd[1]), 
                linewidth = linewidth,
                aes(colour = "word initial")) +
  stat_function(fun = dnorm, n = 101, args = list(mean = sent_mean, sd = sd[2]), 
                linewidth = linewidth,
                aes(colour = "sentence initial"), show.legend = F) +
  geom_area(stat = "function", fun = dnorm, aes(fill = word_lbl), alpha = .25, 
            args = list(mean = word_mean, sd = sd[1])) +
  geom_area(stat = "function", fun = dnorm, aes(fill = sent_lbl), alpha = .25, 
            args = list(mean = sent_mean, sd = sd[2])) +
  geom_curve(aes(x = word_mean, xend = sent_mean, y = 0.01, yend = 0.01),
             curvature = -0.2, 
             size = .25,
             arrow = arrow(length = unit(0.3, "cm"))) +
  annotate("text", x=(sent_mean+word_mean)/2, y=.0115, label=paste0("beta[diff]"), size = 4, parse=TRUE) +
  scale_color_manual(breaks = c(word_lbl, sent_lbl), values = ggthemes::colorblind_pal()(2)) +
  scale_fill_manual(breaks = c(word_lbl, sent_lbl), values = ggthemes::colorblind_pal()(2)) +
  scale_y_continuous(limits = c(0, .012)) +
  scale_x_continuous(labels = scales::comma, limits = c(650, 1150)) +
  labs(x = "Interkey intervals [in ms]", 
       colour = "Text location",
       fill = "Text location") +
  theme(axis.text.y = element_blank(),
        axis.title.y = element_blank(),
        axis.ticks.y = element_blank())
```
</div>

@@

$$
\text{iki}_i \sim\text{ } \text{log}\mathcal{N}(\beta_\text{textlocation[i]}, \sigma_{e}^2)
$$

@@

## Parallel view can be implemented as two distributions mixture-process [@roeser2021modelling]

- Executing motor movements [100-150 msecs, @conijn2019understanding;@waes2019].
- If upstream processes provide output more slowly, then interkey intervals are determined by time taken to complete upstream processes and not motor movement.

@@

```{r fig.width=5}
ggplot(data = data, aes(x)) +
  stat_function(fun = dnorm, n = 101, args = list(mean = word_mean, sd = sd[1]), 
                linewidth = linewidth,
                aes(colour = "word initial"), show.legend = F) +
  geom_area(stat = "function", fun = dnorm, aes(fill = word_lbl), alpha = .25, 
            args = list(mean = word_mean, sd = sd[1])) +
  scale_color_manual(breaks = c(word_lbl, sent_lbl), values = ggthemes::colorblind_pal()(2)) +
  scale_fill_manual(breaks = c(word_lbl, sent_lbl), values = ggthemes::colorblind_pal()(2)) +
  scale_y_continuous(limits = c(0, .012)) +
  scale_x_continuous(labels = scales::comma, limits = c(650, 1150)) +
  labs(x = "Interkey intervals [in ms]", 
       colour = "Text location",
       fill = "Text location") +
  theme(axis.text.y = element_blank(),
        axis.title.y = element_blank(),
        axis.ticks.y = element_blank())

```

@@

$$
\text{iki}_{i} \sim\text{ } \theta \times \text{log}\mathcal{N}(\beta, \sigma_{e}^2) \\
\theta=1
$$

@@

- Fluent typing speed: $\beta$
- Hesitation slowdown (pause duration): $\delta$
- Hesitation probability (pause frequency): $\theta$

@@

### 2024-10-26 slides/sections/textlocations.Rmd
After review update

```{r keyloc, results='asis'}
tibble(`Transition type` = c("Within word", "Below word", "Before sentence"),
                Description = c("Transitions between any letter",
                                "Keypress after space followed by any letter",
                                "Keypress following a space preceding any letter"),
                Example = c("T$^{\\wedge}$h$^{\\wedge}$e c$^{\\wedge}$a$^{\\wedge}$t m$^{\\wedge}$e$^{\\wedge}$o$^{\\wedge}$w$^{\\wedge}$e$^{\\wedge}$d. T$^{\\wedge}$h$^{\\wedge}$a$^{\\wedge}$t[bsp][bsp]e$^{\\wedge}$n i$^{\\wedge}$t s$^{\\wedge}$l$^{\\wedge}$e$^{\\wedge}$p$^{\\wedge}$t.", 
                            "The $^{\\wedge}$cat $^{\\wedge}$meowed. That[bsp][bsp]en $^{\\wedge}$it $^{\\wedge}$slept.", 
                            "The cat meowed. $^{\\wedge}$That[bsp][bsp]en it slept.")) %>% 
  kable() %>% 
  kable_styling(bootstrap_options = c("striped", "hover", "responsive"),
                full_width = T, 
                position = "right",
                font_size = 16) %>% 
  add_footnote("Note: Key intervals that terminated in a revision were removed.") %>% 
  row_spec(0:3, extra_css = "line-height: 3.25;")  # Adjust the number for more spacing

>>>>

```{r keyloc, results='asis'}
tibble(Location = c("Within word", "Below word", "Before sentence"),
                Description = c("Transitions between any letter",
                                "Keypress after space followed by any letter",
                                "Keypress following a space preceding any letter"),
                Example = c("T$^{\\wedge}$h$^{\\wedge}$e c$^{\\wedge}$a$^{\\wedge}$t m$^{\\wedge}$e$^{\\wedge}$o$^{\\wedge}$w$^{\\wedge}$e$^{\\wedge}$d. T$^{\\wedge}$h$^{\\wedge}$e$^{\\wedge}$n i$^{\\wedge}$t s$^{\\wedge}$l$^{\\wedge}$e$^{\\wedge}$p$^{\\wedge}$t.", 
                            "The $^{\\wedge}$cat $^{\\wedge}$meowed. Then $^{\\wedge}$it $^{\\wedge}$slept.", 
                            "The cat meowed. $^{\\wedge}$Then it slept.")) %>% 
  select(-Description) %>% 
  kable() %>% 
  kable_styling(bootstrap_options = c("striped", "hover", "responsive"),
                full_width = T, 
                position = "right",
                font_size = 16) %>% 
  add_footnote("Note: Key intervals that terminated in a revision were removed.") %>% 
  row_spec(0:3, extra_css = "line-height: 3.25;")

### 2024-10-26 tutorial/tutorial.Rmd
After review update

```{r eval=F, engine='Rcpp'}
// Do not run
// Data chunk of Stan code
data {
  int<lower=1> N;                    // number of observations
  int<lower=1> nS;                   // number of ppts
  int<lower=1, upper=nS> ppt[N];     // ppts identifier
  int<lower=1> K;                    // number of conditions
  int<lower=1> condition[N];         // condition identifiers
  vector[N] y;                       // outcome
}
```

@@

```{r}
posterior_diffs <- posterior_in_msecs %>%
  select(-idx) %>% 
  pivot_longer(delta:theta, names_to = "parameter") %>%
  mutate(id = row_number(), .by = location) %>%
  pivot_wider(names_from = location, values_from = value) %>%
  select(-id) %>% # drop id column
  transmute(parameter = str_c(parameter, "[diff]"),
            diff_sent = `before sentence` - `before word`, # compare sentences to word
            diff_word = `before word` - `within word`)     # compare within to before word

@@

and then apply the `BF` function from above first to the difference between before-sentence and before-word transitions. For interpretation it is important to remember that $\theta$ is a probability (which we transformed from logit to probability above); hence the difference between two probabilities can only range between -1 and 1 (in the most extreme case). In other words, a prior SD of 1 is saying anything goes (which is not helpful). Similar to the calculations of the ROPE above, it is useful to contemplate which prior distribution appropriately reflects the null hypothesis for the scale the difference is on (e.g. logit, probability, log msecs, msecs). SDs of 1 and .5 render very weakly informative prior distributions (flat, really) as they assign about the same prior probability to differences across the board. An SD of .1 is a more realistic representation for the distribution of null differences.^[For `brms` fits the `hypothesis` function calculates the BFs indicated as evidence ratio in the output. For a model such as `y ~ 1 + condition`, `hypothesis` can calculate the evidence for the hypothesis that the effect for `condition` is, for example, equal to 0. `hypothesis` can then be applied to the model with a statement that describes the hypothesis that should be evaluated.

```{r eval = F}
hypothesis(brm_model, "conditionb = 0")
```

To calculate the evidence ratio the argument `sample_prior = TRUE` (or `sample_prior = 'yes'`) needs to be specified in `brm`.]

>>>>

```{r eval=F, engine='Rcpp'}
// Do not run
// Data chunk of Stan code
data {
	int<lower=1> N;               // Number of observations
  int<lower=1> nS;              // Number of subjects
  int<lower=1, upper=nS> ppt[N];// Participant identifiers
  int<lower=1> K;               // Number of conditions
  int<lower=1> condition[N];    // Condition identifiers
  vector[N] y;                  // Outcome: interkey intervals
}
```

@@

```{r}
posterior_diffs <- posterior_in_msecs %>%
  select(-idx) %>% 
  pivot_longer(delta:theta, names_to = "parameter") %>%
  mutate(id = row_number(), .by = location) %>%
  pivot_wider(names_from = location, values_from = value) %>%
  transmute(parameter = str_c(parameter, "[diff]"),
            diff_sent = `before sentence` - `before word`, # compare sentences to word
            diff_word = `before word` - `within word`)     # compare within to before word

@@

and then apply the `BF` function from above first to the difference between before-sentence and before-word transitions. For interpretation it is important to remember that $\theta$ is a probability (which we transformed from logit to probability above mostly for ease of explanation; in practice it is a better idea to work with the logit scaled values); hence the difference between two probabilities can only range between -1 and 1 (in the most extreme case). In other words, a prior SD of 1 is saying anything goes (which is not helpful). Similar to the calculations of the ROPE above, it is useful to contemplate which prior distribution appropriately reflects the null hypothesis for the scale the difference is on (e.g. logit, probability, log msecs, msecs). SDs of 1 and .5 render very weakly informative prior distributions (flat, really) as they assign about the same prior probability to differences across the board. An SD of .1 is a more realistic representation for the distribution of null differences.^[For `brms` fits the `hypothesis` function calculates the BFs indicated as evidence ratio in the output. For a model such as `y ~ 1 + condition`, `hypothesis` can calculate the evidence for the hypothesis that the effect for `condition` is, for example, equal to 0. `hypothesis` can then be applied to the model with a statement that describes the hypothesis that should be evaluated, for example `hypothesis(brm_model, "conditionb = 0")` for a null hypothesis. To calculate the evidence ratio the argument `sample_prior = TRUE` (or `sample_prior = 'yes'`) needs to be specified in `brm`.]
